---
title: "schnellerGP: Installation"
author: Wesley S. Burr^1^
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{schnellerGP: Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

^1^ Department of Mathematics, Trent University, Peterborough, ON, Canada

# Introduction

The schnellerGP package contains interfaces to two source libraries (in C
and C++): H2Lib and HODLR. In addition, the package assumes access to 
BLAS, LAPACK, and other standard code libraries. As such, the installation
can be a bit tricky, especially if you haven't previously set up an optimized
BLAS library for your personal system. This vignette will go through
the process of configuring your environment for the installation of the
package, for five different operating system frameworks: Windows 7, Windows 10,
Ubuntu 20.04, Fedora 35, and Mac OSX. Please note that the installation has
**not** been tested on the new M1 Mac chipset, and will not be until one of the
authors gains access to such a machine.


# Windows 7

For Windows, because packages are not as easily installed as, say, a Linux operating
system, the instructions are somewhat more complex. As with the others, if you would
like to optimize a BLAS installation, please refer to the Optimization Vignette.
Note that these instructions are only for installing the **latest** version of
schnellerGP from **source** - compiled Windows binaries can gloss over most of these
issues. Unfortunately, this takes work, and we won't always have the latest versions
available. 

Start by ensuring you have the latest version of **R tools** installed for your version
of R. Testing has been done against R 4.1 and later, but not on R 3.5 or 3.6.
This can be found on [https://cran.r-project.org/bin/windows/Rtools/](CRAN),
as with the R interpreter and interface. Generally speaking, Rtools provides a
miniaturized Linux environment and compiler-linker chain for Windows, based on a 
system called MinGW. Because of this, once you install Rtools there will be a link in your
Start Menu under Rtools 4.0 for **Rtools MinGW 64-bit**. If you click this, you will get
a terminal in the environment that R uses for compiling packages from source. 

There is a note in the Rtools installation guidelines on CRAN, which states that


    After installation is complete, you need to perform one more step to be able to compile R 
    packages: you need to put the location of the Rtools make utilities (bash, make, etc) on the 
    PATH. The easiest way to do so is create a text file .Renviron in your Documents folder which 
    contains the following line:
     
    PATH="${RTOOLS40_HOME}\usr\bin;${PATH}"
    
    If you then restart R, you can verify that make can be found, which should show the path to 
    your Rtools installation:

    ```{r, eval = FALSE}
    Sys.which("make")
    ```
 
Make sure that you name the file **.Renviron** exactly, and not something like **.Renviron.txt**,
which will not work. Since you're already in R to test the above command, also install the
required packages:

```{r, eval = FALSE}
install.packages(c("Rcpp", "RcppEigen", "RcppArmadillo", "knitr", "rmarkdown"))
```

Now, we do the tricky bits, starting with setting up the build environment.
    
## Setting up the pacman / mingw world

If you run Rtools MinGW 64-bit from your start menu, you can then use the **pacman** package manager
to organize system libraries that can be used to compile against. Start with updating the
indexes (package databases):

```bash
pacman -Sy
```

Then, we need the same dependencies that we would compile against for Linux: netCDF, BLAS,
and Cairo. Pandoc we will get by installing rmarkdown, which brings it along on Windows.
We will install cairo, BLAS and netcdf, and for both the 32-bit and 64-bit architectures
(because Windows is weird, and sometimes will just randomly decide to install the i386 version
of a package). The glib2 is required for Cairo to work automatically in the configure.win
call.

```bash
pacman -S mingw-w64-x86_64-netcdf mingw-w64-i686-netcdf
pacman -S mingw-w64-i686-cairo mingw-w64-x86_64-cairo
pacman -S mingw-w64-i686-openblas mingw-w64-x86_64-openblas
pacman -S mingw-w64-i686-glib2 mingw-w64-x86_64-glib2
```

That should set things up (ignoring the customized BLAS - for that, see the Optimization Vignette)
so you can compile. Return to R, and on the prompt run:

```{r, eval = FALSE}
install.packages("schnellerGP_0.1.0.tar.gz", type = "source")
```

(assuming you have placed the .tar.gz file in the current working directory). Alternatively,
use the menu system and run Packages -> Install package(s) from local files ..., then
point the interface at the .tar.gz file and allow it to proceed. If the package compiles
successfully, you can then load it in the usual way, and proceed with your modeling!

# Windows 10

# Ubuntu Desktop 20.04

# Fedora Desktop 35

Assuming you have a working installation of Fedora (tested on 34 and 35), you should have installed
a few background applications:

```bash
sudo dnf install R netcdf netcdf-devel blas blas-devel cairo-devel pandoc
```

In addition, the package depends on several R dependencies, which will be automatically taken
care of if you install via, e.g., RStudio's system. If you prefer to do things manually, 
you will need the following (knitr and rmarkdown are to allow the vignettes to be installed):

```{r, eval = FALSE}
install.packages(c("Rcpp", "RcppEigen", "RcppArmadillo", "knitr", "rmarkdown"))
```

Note that setting up alternative BLAS options for Linux is fairly straightforward:
see the Optimization Vignette for more on how to do this for Fedora.

Once you have ensured the dependencies are available, the installation is straightforward. 
From within R, you can run:

```{r, eval = FALSE}
install.packages("schnellerGP_0.1.0.tar.gz", type = "source")
```
or from the command line,
```bash

```
# Mac OSX 10.??



