---
title: "schnellerGP: Installation and Performance"
author: Wesley S. Burr^1^
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{schnellerGP: Installation and Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

^1^ Department of Mathematics, Trent University, Peterborough, ON, Canada

# Introduction

The schnellerGP package contains interfaces to a source library (in
 C++): HODLRlib, which "is a library consisting of fast matrix 
 operations for matrices based on the Hierarchical Off-Diagonal Low-Rank (HODLR) structure."
As such, the package works best when given access to OpenMP and a good BLAS (i.e., not
the default shipping) library. As such, the installation for best performance
can be quite a bit trickier than a standard R package, as while the 
install.packages() step **will** install the package, it will not perform to 
expectations without more work.

This vignette goes through
the process of configuring your environment for the installation of the
package, for five different operating system frameworks: Windows 7, Windows 10,
Ubuntu 20.04, Fedora 35, and Mac OSX, with that last discussed for both Intel
and M1 (Mac) chipsets.

# Windows 10

**Note**: Using Windows with a) compiled code; b) high performance requirements;
c) linear algebra is basically a pain. 

For Windows, because packages are not as easily installed as, say, a Linux operating
system, the instructions are somewhat more complex.
Note that these instructions are only for installing the **latest** version of
schnellerGP from **source** - compiled Windows binaries can gloss over most of these
issues. Unfortunately, this takes work, and we won't always have the latest versions
available. Note that no significant differences have been observed for installing
this package on either Windows 7 or 10; Windows 11 has not been tested at this time.

Start by ensuring you have the latest version of **R tools** installed for your version
of R. Testing has been done against R 4.1 and later, but not on R 3.5 or 3.6.
This can be found on [https://cran.r-project.org/bin/windows/Rtools/](CRAN),
as with the R interpreter and interface. Generally speaking, Rtools provides a
miniaturized Linux environment and compiler-linker chain for Windows, based on a 
system called MinGW. Because of this, once you install Rtools there will be a link in your
Start Menu under Rtools 4.0 for **Rtools MinGW 64-bit**. If you click this, you will get
a terminal in the environment that R uses for compiling packages from source. 

There is a note in the Rtools installation guidelines on CRAN, which states that

    After installation is complete, you need to perform one more step to be able to compile R 
    packages: you need to put the location of the Rtools make utilities (bash, make, etc) on the 
    PATH. The easiest way to do so is create a text file .Renviron in your Documents folder which 
    contains the following line:
     
    PATH="${RTOOLS40_HOME}\usr\bin;${PATH}"
    
    If you then restart R, you can verify that make can be found, which should show the path to 
    your Rtools installation:

    ```{r, eval = FALSE}
    Sys.which("make")
    ```
 
Make sure that you name the file **.Renviron** exactly, and not something like **.Renviron.txt**,
which will not work. Since you're already in R to test the above command, also install the
required packages:

```{r, eval = FALSE}
install.packages(c("Rcpp", "RcppEigen", "RcppArmadillo", "knitr", "rmarkdown"))
```

## Optimized BLAS

The precompiled R distribution that you likely downloaded and installed on your Windows
machine is called the **reference BLAS**, and is designed to be very, very stable and cross-platform
compatible, but not particularly optimized for performance. Significant improvements in almost
all linear algebra operations can be observed by installing more optimized libraries. Given the
use of HODLRlib, which is optimized for Intel Math Kernel Library (MKL), if you are planning on 
running any serious code with this package on Windows, you really need to update your BLAS. However,
doing this on Windows is **really** hard. The Intel MKL compilers do not match with those needed
to compile R from source, so there is no easy way to link in the MKL without some shenanigans.

Before you begin, you can use the **sessionInfo** in R to check which version you are running:
```{r, eval = FALSE}
sessionInfo()
```
You will see something like

    Matrix products: default

if you are running the default BLAS library, and do not have a custom LAPACK installed. This indicates
that you will benefit from the change suggested here. It is probably worth grabbing a benchmark
script before you start, to verify that the changes have actually given some speedup. One
[https://mac.r-project.org/benchmarks/](set of good examples) is from Simon Urbanek's old
site, even if they are a bit dated. Using Option 2 or 3 below results in a minimum of a 5x speedup
on many operations; the increases can be even more for certain classes of matrix operations.

### Option 1: Install MKL, Manually Move Libraries

The Intel MKL library is included as part of the [https://www.intel.com/content/www/us/en/developer/tools/oneapi/onemkl.html](Intel oneAPI Base Toolkit),
which is freely available for download. This framework included BLAS and LAPACK operations,
sparse linear algebra functions, Fast Fourier Transforms, and other functions which to some extent
may be used by R to do statistical modeling. If you go to the provided link, and choose the Offline
Installer, it is 3.5GB in size compressed (and 26GB fully installed), but can be downloaded once and installed, and then (likely) never
touched again. The **Online** version is only 13.4MB, but then will download what you need dynamically
after installation. This is still multiple gigabytes of space (6.8GB if you only select the
MKL and nothing else).

Download the oneAPI Base Toolkit, and install at least the MKL on your computer. Then, navigate
to your R program directory, typically in 

    C:\Program Files\R\R-4.1.2\bin\x64

and copy all of the contents of the mkl folder, typically in

    C:\Program Files (x86)\Intel\oneAPI\mkl\latest\redist\intel64\

to your R bin\x64 directory. Within the latter, find the files named **Rblas.dll** and **Rlapack.dll** and rename
them to **Rblas.bak** and **Rlapack.bak**. Note that you will need file extensions visible in your Windows Explorer
to be able to do this properly (click **View** then check the box for File name extensions). Now, take the file 
**mkl_rt.dll** which should have copied over along with the rest of the Intel files, and copy it twice, still
within the bin\x64 directory. Rename the first to **Rblas.dll**, and the second to **Rlapack.dll**.
Restart R or RStudio if you have it open, and when you run **sessionInfo()**, your information
should have updated. 

**NOTE**: this doesn't currently work well. The oneAPI is a bit different to the original MKL installer,
for which this was a good drop-in solution.

### Option 2: Microsoft R Open (Now Deprecated)

The **easiest** way to get improved performance in R on Windows **was** to simply install 
[https://mran.microsoft.com/rro](Microsoft R Open). This was an open source fork of R 
(currently 4.0.2 as of January, 2022) which had been optimized to use the Intel MKL and other, related
performance libraries. It also, less fortunately, comes with a customized, snapshotted CRAN, which
while fast, is not updated in between releases of Microsoft R Open. In addition, this program has
been deprecated as of June 2021, and will no longer be updated moving forward, making it a poor
overall choice. However, for now, R 4.0.2 isn't **that** old, so this is a nice easy fix if you just
need some speed for now.

All you need to do is run [https://www.rstudio.com/products/rstudio/download/](RStudio),
install Microsoft R Open, and then switch which R you are using within RStudio when you want to work
on linear algebra applications. In RStudio, under **Global Options, General** you can specify which R version you
wish to use - if you restart RStudio after installing Microsoft R Open, it will appear as one of your
options. You may find it suitable for everyday use, in which case you won't be able to easily
update your CRAN packages to the latest and greatest, but you will have extremely fast linear algebra
libraries on the back end. Note that if you run
```{r, eval = FALSE}
sessionInfo()
```
you will not see any particular information about your BLAS library, but if you are referencing
the R version provided by Microsoft R Open (4.0.2, currently), it will be using the correct,
higher performance libraries.

### Option 3: Microsoft R Open, Just the Libraries

Repeat option 2, install Microsoft R Open, even if deprecated, then take three files:

    C:\Program Files\Microsoft\R Open\R-4.0.2\bin\x64\libiomp5md.dll
    C:\Program Files\Microsoft\R Open\R-4.0.2\bin\x64\Rblas.dll
    C:\Program Files\Microsoft\R Open\R-4.0.2\bin\x64\Rlapack.dll

and copy them to your 

    C:\Program Files\R\R-4.1.2\bin\x64\

directory. This does violate the Microsoft R Services MKL license, as it separates the MKL
out of Microsoft R Open, but it's a solution which gets you the speed if you also need to
continue using Microsoft R Open past its deprecation, and can't (or won't) install the 
MKL directly.

## Install schnellerGP from Source

Now that you have an updated BLAS library installed, proceed to install **schnellerGP**
by installing from the source tar.gz file - either type out the following
(assuming your download is in the current working directory), or use the Install Packages
interface in RStudio to point to the correct file. 

```{r, eval = FALSE}
install.packages("schnellerGP_0.1.0.tar.gz", type = "source", repo = NULL)
```

If the package compiles
successfully, you can then load it in the usual way, and proceed with your modeling!

# Windows 7

The instructions for Windows 7 are the same as Windows 10, except you cannot install Microsoft R
Open, as it does not support Windows 7 any more. Thus, while you can install Rtools, and install
**schnellerGP**, it will not perform well. We strongly encourage the upgrade to a more recent
operating system (Windows 10 is 7 years old as of 2022; Windows 7 is 13 years old!).

# Ubuntu Desktop 20.04

# Fedora Desktop 35

Assuming you have a working installation of Fedora (tested on 34 and 35), you should have installed
a few background applications:

## Optimized BLAS

```bash
sudo dnf install R blas blas-devel R-flexiblas pandoc
dnf config-manager --add-repo https://yum.repos.intel.com/mkl/setup/intel-mkl.repo
rpm --import https://yum.repos.intel.com/intel-gpg-keys/GPG-PUB-KEY-INTEL-SW-PRODUCTS-2019.PUB
dnf install intel-mkl
```

Once this is done, if you launch an R session, you should be able to dynamically change your
BLAS library at will. For example,

```{r, eval = FALSE}
library(flexiblas)
flexiblas_load_backend("/opt/intel/mkl/lib/intel64/libmkl_rt.so")
flexiblas_list_loaded()
```
[1] "OPENBLAS-OPENMP"                        
[2] "/opt/intel/mkl/lib/intel64/libmkl\_rt.so"

Note that OpenBLAS is still a great package, and for many operations, is actually faster than
Intel's MKL (e.g., matrix products). However, as HODLRlib works best with MKL, this is why
we recommend it be present.

In addition, the **schnellerGP** package depends on several R dependencies.
You will need the following (knitr and rmarkdown are to allow the vignettes to be installed):

```{r, eval = FALSE}
install.packages(c("Rcpp", "RcppEigen", "RcppArmadillo", "knitr", "rmarkdown"))
```

Once you have ensured the dependencies are available, the installation is straightforward. 
From within R, you can run:

```{r, eval = FALSE}
install.packages("schnellerGP_0.1.1.tar.gz", type = "source", repo = NULL)
```
or from the command line,
```bash
R CMD INSTALL schnellerGP_0.1.1.tar.gz
```

# Mac OSX 12.0 - Intel Chipset

# Mac OSX 12.0 - Apple (ARM) Chipset (M1)

New Mac systems, as of 2020, come with the new Apple M1 chip variants, which are custom
Apple silicon based on the ARM architecture. This, of course, complicates our lives. 

To start, you should ensure you have:

* installed XCode 13.2
* installed XQuartz 2.8.1 (required for R)
* installed R 4.1 or later, as the appropriate arm64 installer (Note: if you install the x86 binary,
  it will work, but will run under Rosetta emulation, and you will lose speed)
* installed RStudio (version 2021.09 or later)
* installed a conda environment (we assume Anaconda3 below)

If you then launch RStudio, it may ask you to install the "command line developer tools",
as an extension of XCode. Do this, and you should then be largely set up to compile packages
from source. 

**Note**: I also install the gfortran compiler (GNU Fortran 11 arm64 fork, as described
on the cran.r-project.org page).

## Optimized BLAS

Intel's MKL obviously isn't going to work on a non-x64 architecture, so our options are
limited to OpenBLAS to get better linear algebra performance. Thankfully, it's not that
hard to set up. Assuming you've done the above, you should be able to
download OpenBLAS (version 0.3-17 as of latest) using **conda**, which maintains packages
for the new M1 chips. In the following, I use the full Anaconda3 environment to install
openblas, and then symlink 

```{bash, eval = FALSE}
conda install openblas
cd /Library/Frameworks/R.framework/Resources/lib
ln -sf /opt/openblas/lib/libopenblas.dylib libRblas.dylib
```

**Note**: the -sf tag on ln makes the link symbolic, and removes the existing destination file.
R keeps a copy of the blas library as libRblas.0.dylib, so we can just overwrite the base
file and the original is still available if we ever want to revert.

Once you've done this, you are ready to install **schnellerGP** from source if you wish.



